Network Data Model and BerlinMOD Benchmark

### 2010-07-05 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Further operations added.

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[ruled]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\usepackage{todonotes}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, the translation
from the \bmodb{} into the network constrained data model, and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\listoftodos
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space (DMFS) and data models
for network constrained moving objects (NCDM). For both categories several different
data models have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis,HERMESMDCPelekis}
for DMFS and \cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,NetworkVazirgiannis}
for NCDM, to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

@@

The most NCDM use edge based graph representations for the representation of the
underlying network data only a small number of NCDM uses route oriented data models
or combine route and edge based data models.
.
\cite{NetworkVazirgiannis} proposes an edge based NCDM. The edges and their attributes
are stored in a relation representing the network as an undirected graph. Moving
objects are assumed to drive always on the path with the lowest cost,
in terms of distance or travel-time. They are defined by the source point, the
target point, and the starting time instant of the trip. The trajectory is
computed by this assumption using the length and speed attributes of the graph
edges within the shortest respectively fastest path computation.
The advantage of this definition is the reduced storage space for the representation
of moving point objects. The drawback is the high computational effort for query
evaluation on moving objects.

@@

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard attribute, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relationships condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in Section \ref{sec:queries}\todo[caption={check reference}]{Check reference when paper finished. Parts of querie explanations might be moved to the appendix.} together with our NDM algorithms for these queries.
\section{Implementation Issues}
\label{sec:implNDM}
We started a description of the implementation of NET in \secondo{} in Section
\ref{sec:netdatamod} with the description of the implementation proposed in
\cite{NetworkGueting}. In the meantime this implementation has been changed and
extended in some points to support faster query execution.

@@

The implementation of the \dt{ugpoint} has been changed to:
\begin{ttfamily}
\begin{itemize}
  \item [] ugpoint: record \{gp1: gpoint; gp2: gpoint; ti : Interval;\}
\end{itemize}
\end{ttfamily}
Where $ti$ is a time interval consisting of two time instants $t1$ and $t2$ and
two Boolean flags, indicating if $t1$ respectively $t2$ is part of the time interval
or not.

@@

We use $m$ (respectively $m_1,\ m_2$) as the number of units of the \dt{mgpoint} value
(resp. first, second \dt{mgpoint} value),
$r$ (resp. $r_1$, $r_2$) as the number of $route\ intervals$ of the \dt{gline} value
or the $trajectory$ attribute of an \dt{mgpoint} value (resp. of the first, second
argument), $R$ as the number of routes in the network, and $p$ as the number of
time intervals in a \dt{periods} value.

\begin{table}[H]
  \caption{Simple Operations on Network Data Types}
  \label{tab:simplenetoperations}
  %\begin{scriptsize}
  \begin{center}
  \begin{tabularx}{1.0\textwidth}{|l|l|X|l|}
  \hline
  \textbf{Name}& \textbf{Signature}& \textbf{Explanation}& \textbf{Complexity}\\
  \hline
\op{val}&\dt{igpoint} $\rightarrow$ \dt{gpoint}& Returns the \dt{gpoint} of the
\dt{igpoint} value&O(1)\\
  \hline
\op{inst}&\dt{igpoint} $\rightarrow$ \dt{instant}& Returns the time instant of the
\dt{igpoint} value &O(1)\\
  \hline
  \op{initial}&\dt{mgpoint} $\rightarrow$ \dt{igpoint}& Returns the first position
and the start time of the \dt{mgpoint} value& O(1)\\
  \hline
  \op{atinstant}&\dt{mgpoint} $\times$ \dt{instant} $\rightarrow$ \dt{igpoint}&
Returns the network position of the \dt{mgpoint} value at the given time instant& O($\log m$)\\
  \hline
  \op{length}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the $length$ of the argument.&O(1)\\
  \hline
\op{isempty}&\dt{gline} $\rightarrow$ \dt{bool}& Returns \true{} if the \dt{gline}
has no $route\ intervals$ & O(1)\\
  \hline
\op{no\_components}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the number of
units of the \dt{mgpoint} &O(1)\\
  \hline
\op{trajectory}&\dt{mgpoint} $\rightarrow$ \dt{gline}& Returns the $trajectory$
of the \dt{mgpoint} value as \dt{gline}&O($r$)\\
  \hline
  \op{units}&\dt{mgpoint} $\rightarrow$ \dt{stream}(\dt{ugpoint})& Returns a stream
of the \dt{ugpoint} values of the \dt{mgpoint} value& O($m$)\\
  \hline
  \op{routes}& \dt{network} $\rightarrow$ \dt{relation} & Returns the $routes$
relation fo the \dt{network} object & O($R$)\\
  \hline
\op{gpoint2rect}& \dt{gpoint} $\rightarrow$ \dt{rectangle} & Returns a $netbox$
for the \dt{gpoint} value & O(1)\\
  \hline
\op{routeintervals}&\dt{gline} $\rightarrow$ \dt{stream}(\dt{rectangle})& Returns
the $netbox$ for each $route\ interval$ of the \dt{gline} value& O($r$)\\
  \hline
  \op{unitbox}& \dt{ugpoint} $\rightarrow$ \dt{rectangle3D} & Returns the $netbox$
of the \dt{ugpoint} & O(1)\\
  \hline
  \end{tabularx}
  \end{center}
  %\end{scriptsize}
\end{table}

The operations used to construct the network object, and to translate spatial and
spatio-temporal values into network values are described in Section \ref{sec:Translation}
in more detail.

For the Euclidean Distance computation we use the operations \op{gline2line} and
\dt{mgpoint2mpoint}, which compute spatial respectively spatio-temporal values
from the network values.

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[ruled]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{todonotes}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data
models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two
representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, the translation
from the \bmodb{} into the network constrained data model, and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\listoftodos
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space (DMFS) and data 
models
for network constrained moving objects (NCDM). For both categories several different
data models have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis
,HERMESMDCPelekis}
for DMFS and 
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,
NetworkVazirgiannis}
for NCDM, to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

@@

The most NCDM use edge based graph representations for the representation of the
underlying network data only a small number of NCDM uses route oriented data models
or combine route and edge based data models.

\cite{NetworkVazirgiannis} proposes an edge based NCDM. The edges and their 
attributes
are stored in a relation representing the network as an undirected graph. Moving
objects are assumed to drive always on the path with the lowest cost,
in terms of distance or travel-time. They are defined by the source point, the
target point, and the starting time instant of the trip. The trajectory is
computed by this assumption using the length and speed attributes of the graph
edges within the shortest respectively fastest path computation.
The advantage of this definition is the reduced storage space for the representation
of moving point objects. The drawback is the high computational effort for query
evaluation on moving objects.

@@

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard attribute, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relationships condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in Section
\ref{sec:queries}\todo[caption={check reference}]{Check reference when paper
finished. Parts of query explanations might be moved to the appendix.}
together with our NDM algorithms for these queries.
\section{Implementation Issues}
\label{sec:implNDM}
We started a description of the implementation of NET in \secondo{} in Section
\ref{sec:netdatamod} with the description of the implementation proposed in
\cite{NetworkGueting}. In the meantime this implementation has been changed and
extended in some points to support faster query execution.

@@

The implementation of the \dt{ugpoint} has been changed to:
\begin{ttfamily}
\begin{itemize}
  \item [] ugpoint: record \{gp1: gpoint; gp2: gpoint; ti : Interval;\}
\end{itemize}
\end{ttfamily}
Where $ti$ is a time interval consisting of two time instants $t_1$ and $t_2$
and
two Boolean flags, indicating if $t_1$ respectively $t_2$ is part of the time
interval
or not.

@@

We use $m$ (respectively $m_1,\ m_2$) as the number of units of the
\dt{mgpoint} value (resp. first, second \dt{mgpoint} value),
$r$ (resp. $r_1$, $r_2$) as the number of $route\ intervals$ of the \dt{gline}
value or the $trajectory$ attribute of an \dt{mgpoint} value (resp. of the
first, second argument), $R$ as the number of routes in the network, and $p$ as
the number of time intervals in a \dt{periods} value.
\begin{table}
  \caption{Simple Operations on Network Data Types}
  \label{tab:simplenetoperations}
  %\begin{scriptsize}
  \begin{center}
  \begin{tabularx}{1.0\textwidth}{|l|l|X|l|}
  \hline
  \textbf{Name}& \textbf{Signature}& \textbf{Explanation}& \textbf{Complexity}\\
  \hline
\op{val}&\dt{igpoint} $\rightarrow$ \dt{gpoint}& Returns the \dt{gpoint} of the
\dt{igpoint} value&O(1)\\
  \hline
\op{inst}&\dt{igpoint} $\rightarrow$ \dt{instant}& Returns the time instant of the
\dt{igpoint} value &O(1)\\
  \hline
  \op{initial}&\dt{mgpoint} $\rightarrow$ \dt{igpoint}& Returns the first position
and the start time of the \dt{mgpoint} value& O(1)\\
  \hline
  \op{atinstant}&\dt{mgpoint} $\times$ \dt{instant} $\rightarrow$ \dt{igpoint}&
Returns the network position of the \dt{mgpoint} value at the given time instant& O($\log m$)\\
  \hline
  \op{length}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the $length$ of the argument.&O(1)\\
  \hline
\op{isempty}&\dt{gline} $\rightarrow$ \dt{bool}& Returns \true{} if the \dt{gline}
has no $route\ intervals$ & O(1)\\
  \hline
\op{no\_components}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the number of
units of the \dt{mgpoint} &O(1)\\
  \hline
\op{trajectory}&\dt{mgpoint} $\rightarrow$ \dt{gline}& Returns the $trajectory$
of the \dt{mgpoint} value as \dt{gline}&O($r$)\\
  \hline
  \op{units}&\dt{mgpoint} $\rightarrow$ \dt{stream}(\dt{ugpoint})& Returns a stream
of the \dt{ugpoint} values of the \dt{mgpoint} value& O($m$)\\
  \hline
  \op{routes}& \dt{network} $\rightarrow$ \dt{relation} & Returns the $routes$
relation fo the \dt{network} object & O($R$)\\
  \hline
\op{gpoint2rect}& \dt{gpoint} $\rightarrow$ \dt{rectangle} & Returns a $netbox$
for the \dt{gpoint} value & O(1)\\
  \hline
\op{routeintervals}&\dt{gline} $\rightarrow$ \dt{stream}(\dt{rectangle})& Returns
the $netbox$ for each $route\ interval$ of the \dt{gline} value& O($r$)\\
  \hline
  \op{unitbox}& \dt{ugpoint} $\rightarrow$ \dt{rectangle3D} & Returns the $netbox$
of the \dt{ugpoint} & O(1)\\
  \hline
  \end{tabularx}
  \end{center}
  %\end{scriptsize}
\end{table}

The operations \op{thenetwork}, \op{mpoint2mgpoint}, \op{line2gline}.
\op{point2gpoint}, and \op{polygpoints} which are used to construct the network
object, and to translate spatial and spatio-temporal values into network values
are described in Section  \ref{sec:Translation} in more detail.

For the Euclidean Distance computation we retranslate our network values into
spatial (\op{gline2line}) respectively spatio-temporal (\op{mgpoint2mpoint})
values.

The operation \op{gline2line} uses the Algorithm \ref{alg:gline2line} to
translate a \dt{gline} value into an \dt{line} value. 
\begin{algorithm}
  \caption{\op{gline2line}($gl$)}
  \label{alg:gline2line}
  \begin{algorithmic}[1]
    \FOR {each $route\ interval$ of $gl$}
      \STATE Use B-Tree index of $routes$ relation to get route curve of the
$route\ interval$
      \STATE Perform binary search on line segments of route curve to find the
      start position of the $route\ interval$
      \STATE Copy line segments to result until end pos of $route\ interval$ is
reached.
    \ENDFOR
    \RETURN resulting line
  \end{algorithmic}
\end{algorithm}
The loop from line 1 to
line 5 is repeated $r$ times. The B-Tree search in line 2 take O($\log{R}$)
time. The binary search in line 3 takes O($\log l_i$) time if the number of line
segments of route curve $Rr_i$ is $l_i$. The copy operation in line 4 takes
O($l_i$) time in the worst case. The return of the result has a worst case
time complexity of O($\sum_{i=1}^{r}{l_i}$). For the whole algorithm we
get
\[O(r \log{R} + \sum_{i=1}^{r}{\log{l_i}} + 2\sum_{i=1}^{r}{l_i}) 
= O(r \log{R} + \sum_{i=1}^{r}{l_i})\].

The operation \op{mgpoint2mpoint} is described by Algorithm
\ref{alg:mgpoint2mpoint}. 
\begin{algorithm}
  \caption{\op{mgpoint2mpoint}($mgp$)}
  \label{alg:mgpoint2mpoint}
  \begin{algorithmic}[1]
    \STATE $actRID$ = -1
    \STATE Initialize resulting $mpoint$
    \FOR{each unit $curUGP$ of $mgp$}
      \IF{not (rid from $curUGP$ == $actRID$)}
        \STATE $actRID$ = rid from $curUGP$
        \STATE $rc$ = route curve of $actRID$ 
        \COMMENT{$rc$ determined using B-Tree Index of $routes$ relation}
        \STATE Perform binary search on the line segments of $rc$ to find
        line segment $l$ with start position of $curUGP$
        \STATE $upstart$ = x,y-coordinates of start position
      \ENDIF
      \IF{end position == start position}
        \STATE add unit $upstart,\ upstart$ to $mpoint$ 
      \ELSE
        \IF{end position is on $l$}
          \STATE $upend$ = x,y-coordinates of end position
          \STATE add unit $upstart,\ upend$ to $mpoint$
          \STATE $upstart$ = $upend$
        \ELSE
          \STATE Follow $rc$ in moving direction of $curUGP$
          \WHILE{not(end position is on $l$)}
            \STATE add unit from $upstart$ to $segment\ end\ position$ to
$mpoint$ 
            \STATE $upstart$ = $segment\ end\ position$
            \STATE $l$ = next segment in moving direction
          \ENDWHILE
          \STATE $upend$ = x,y-coordinates of end position
          \STATE add unit from $upstart$ to $upend$ to $mpoint$
        \ENDIF
      \ENDIF
    \ENDFOR
    \RETURN $mpoint$
  \end{algorithmic}
\end{algorithm}
The for-loop from line 3 to line 28 is repeated $m$ times. The statements from
line 5 to line 8 take O($\log R + l_i$) every time they are executed. But they
are only executed at the first run of the loop and if the car changes the route.
We have three different cases for the computing of the $mpoint$ units. The
simple cases from line 10-11 and line 13-16 have a time complexity of O(1).
The third case has a worst case complexity from O($l_i$) if $l_i$ is the number
of line segments of the current route curve. The last line 29 has a time
complexity of O($m_p$) im $m_p$ is the number of units of the resulting
\dt{mpoint} value. In the worst case we get a time complexity of O($m \log {R}
+ \sum_{i=1}^{m}{l_i})
